Are translation equivalents special? Evidence from simulations and empirical data from bilingual infants

### 2021-06-18 paper/paper_TE_bilingual_vocabulary_model.Rmd
fixed wordings

author: 
  - name          : "Rachel Ka-Ying Tsui"
    affiliation   : "1"
    corresponding : yes    
    address       : "Department of Psychology, 7141 Sherbrooke St. West, Montreal, QC, Canada, H2T1V2.  Current address: Laboratory for Language Development, RIKEN Center for Brain Science, 2-1 Hirosawa, Wako-shi, Saitama, Japan, 351-0198"
    email         : "rachelkytsui@gmail.com"
  - name          : "Ana Maria Gonzalez-Barrero"
    affiliation   : "1" 
  - name          : "Esther Schott"
    affiliation   : "1" 
  - name          : "Krista Byers-Heinlein"
    affiliation   : "1"

@@

Bilingual children must learn words that take a different form in each of their languages, but share the same or highly similar meanings. For instance, to refer to the same crisp red-skinned fruit, an English–French bilingual child must use the word “apple” when speaking English, and the word _“pomme”_ when speaking French. These cross-language synonyms are known as translation equivalents (also called doublets; Umbel et al., 1992), and are observed amongst bilingual children’s first words (e.g., David & Wei, 2008; De Houwer et al., 2006; Pearson et al., 1995). Translation equivalents are thought to hold a special status in a bilingual’s developing lexicon due to the strong overlap in their semantics. For example, studies with bilingual toddlers show that the associative semantic properties of a word in one language facilitate the activation of its translation equivalent (e.g., Bilson et al., 2015; Floccia et al., 2020; Jardak & Byers-Heinlein, 2019). That is, upon hearing the English word “apple”, the corresponding French word _“pomme”_ is more easily activated in bilinguals’ minds. In vocabulary acquisition, bilingual children must learn a first label for a referent (a “singlet”; Umbel et al., 1992) before they can learn its translation equivalent. Is translation equivalent learning different from singlet learning? The current paper contrasts three competing accounts: 1) translation equivalents are harder to learn than singlets (Avoidance Account), 2) translation equivalents are easier to learn than singlets (Preference Account), and 3) translation equivalents are similar to learn than singlets (Neutral Account). To adjudicate between these accounts, we introduce the Bilingual Vocabulary Model, which provides a computational account of vocabulary learning, with parameters including bilinguals’ vocabulary in each language and their developmental level. In Study 1, we use the Bilingual Vocabulary Model to derive a set of predictions, which we then test against vocabulary data from 200 18- to 33-month-old bilingual children in Study 2.

@@

Other evidence suggesting translation equivalents might be easier to learn than singlets comes from a study that included vocabulary-checklist data from 254 monolingual and 181 bilingual children aged 6 months to 7.5 years (Bilson et al., 2015). The researchers used a network analysis approach to investigate how translation equivalents are learned, focusing on the semantic relationships between the words (e.g., words like “cat” and “dog” are strongly semantically related). Using a statistical model that allowed free semantic relations among vocabulary data from monolingual and bilingual children, the results suggested that words were learned faster when they were semantically connected to more known words in children’s lexicons. This effect applied not only to words within the same language, but also to words across languages including translation equivalents (e.g., English “dog” and French _“chien”_) and words that had other cross-language relations (e.g., “cat” and _“chien”_). The authors then simulated bilingual vocabulaires by modeling bilingual lexicons as combinations of two independent vocabulary-size-matched monolinguals. Comparison with actual bilingual children’s vocabulary data revealed that bilingual children acquired more translation equivalents than predicted by the simulation. The authors therefore concluded that bilingual children learn translation equivalents more easily than singlets. Note that, in their study, expected translation equivalent knowledge was simulated based on the number of lexical items that overlapped between two randomly chosen English monolinguals (e.g., whether both monolinguals knew the word “cat”). However, it is unclear whether this is an appropriate point of comparison for bilingual children as this approach may overlook variables that impact bilinguals’ vocabulary learning including vocabulary size in each language and the developmental level of a child — a point that we will return to later in the introduction.

@@

### Neutral Account: Translation equivalents are similar to learn than singlets
The previous accounts rely on the idea that bilingual vocabulary development unfolds differently than monolingual development, as monolinguals encounter only singlets but bilinguals encounter both singlets and translation equivalents. There is an underlying assumption that translation equivalent learning is somehow special relative to singlet learning — the Avoidance Account proposes that translation equivalents are harder to learn than singlets, whereas the Preference Account proposes that translation equivalents are easier to learn than singlets. However, it is also possible that translation equivalents are neither harder nor easier for bilingual children to learn than singlets. We call this the Neutral Account.

The Neutral Account implies that bilingual children’s two languages develop relatively independently. Indeed, language and processing measures for bilingual children tend to be tightly correlated within a particular language, with weakly if at all correlated across languages. For example, 30-month-old bilingual children’s processing efficiency in a particular language closely correlated with vocabulary size in that language, but was unrelated to vocabulary size in their other language (Marchman et al., 2010). Due to differences in the amount of language exposure, bilingual children seldom show equal vocabulary growth in both of their languages (e.g., Pearson & Fernandez, 1994; Pearson et al., 1997), and the amount of exposure to a particular language has been reported to modulate the within-language association between language processing ability and vocabulary size (Hurtado et al., 2014). Bilingual children with greater exposure to a particular language tended to process that language faster, and in turn learned more words in that language.

@@

### Vocabulary size in bilinguals’ two languages
Because translation equivalents are words from different languages that refer to the same concept, the number of words a bilingual knows in each of their languages will necessarily constrain the number of translation equivalent pairs they could possibly know. For example, a child with a less balanced vocabulary across the two languages might only say 5 words in one language but many more words in the other language; this means that the child could only produce a maximum of 5 translation equivalents, regardless of how many words they know in their other language. Conversely, it seems reasonable to expect that if a child knows a similar number of words in each language and thus has a more balanced vocabulary across the two languages, there would be more potential for some of those words to be translation equivalents.

Balance between the two vocabulary sizes is a function of the number of words bilingual children produce in each language, which tends to be tightly linked to their exposure to each language. In general, more language exposure leads to larger vocabulary size (e.g., Barnes & Garcia, 2013; Boyce et al., 2013; Hurtado et al., 2014; Marchman et al., 2010; Pearson et al., 1997; Place & Hoff, 2011). Bilingual children usually know more words in the language in which they have greater exposure (i.e; dominant language) relative to the language in which they have less exposure (i.e., non-dominant language; Pearson et al., 1997; Place & Hoff, 2011). This is because the more often a bilingual hears a language, the more opportunities there will be for learning new words in that language.

@@

\newpage
```{r box1, results="asis", tab.cap = NULL}
box1 <- data.frame(title = "To test the independence of two events from the same population, as an example, we might ask whether Psychology majors are more likely to be left-handed. To determine independence, we must know the probability of being a psychology major, the probability of being left handed, and the probability of being both a psychology major and left-handed. The numerator of these probabilities will be the number of students who are left-handed, psychology majors, and both (respectively) and the denominator will be the total population of students we are observing. Imagine a college of 1000 students. If 100/1000 (or 1/10) students are left-handed, and 200/1000 (or 1/5) students are psychology majors, then if these variables are independent we expect a proportion of 1/10 × 1/5 = 1/50 students to be left-handed psychology majors. To determine the number we expect to observe in the college, we multiple 1/50 × 1000 = 20 students. When we compare this expected number to the actual number of left-handed students, there are three possible outcomes. First, we may observe many more than 20 left-handed psychology students at the college (say 100 students), which suggests that being left-handed increases the probability of majoring in psychology. Or, second, we may observe many fewer than 20 left-handed psychology students at the college (say 5 students), this suggests that being left-handed decreases the probability of majoring in psychology (in this example by a factor of 1/4). Finally, if left-handedness and majoring in psychology are independent, we can predict the number of left-handed psychology students by multiplying the observed number of left-handed students (100) by the observed number of psychology students (200), and dividing by the total population of the college (1000), so for example 100×200/1000 = 20. Thus, comparing expected and observed numbers can inform us about the independence of the underlying phenomena.") %>%
  rename("Box 1. Example of test for independence." = title)

@@

The model takes four main parameters: the number of words produced in the dominant language (DOM), the number of words produced in the non-dominant language (NONDOM), vocabulary size of potentially learnable words in each language (LEARNABLE), and a bias parameter (BIAS) which indicates whether the model is biased towards (BIAS > 1) or against (BIAS < 1) learning translation equivalents. The language in which a child knows more words is the dominant language, whereas the one in which a child knows fewer words is the non-dominant language. 
Next, we turn to the LEARNABLE parameter (i.e., the number of potentially learnable words). If DOM and NONDOM are measured with an instrument such as the MacArthur-Bates Communicative Development inventories (CDI; Fenson et al., 2007), one option would be to set LEARNABLE to be the total number of items on the CDI. For convenience, consider the effect of setting LEARNABLE to 600, as a round number (the actual number of CDI items is usually slightly higher than 600, depending on the language of the adaptation). Very young children would not be expected to know many of the “harder” words on the CDI, such as “lawn mower”, “sidewalk”, or “vitamins'', due to children’s immature cognitive machinery and conceptual development. A more reasonable solution might be to determine how many CDI words are potentially learnable given the child’s developmental level, which could be approximated by their age. For example, imagine that Jamie who is 18 months old produces 50 English words and 20 French words, thus a total of 70 words. Monolingual children his age with the very largest productive vocabularies (those at the 90th percentile averaging between English and French norms) produce a total of 245 words (retrieved from the Wordbank database; Frank et al., 2016). Although there is likely considerable individual variability as to the cognitive capacity even amongst children of the same age, we argue that this provides a reasonable — if imperfect — estimate of the number of learnable words (LEARNABLE) that a child of Jamie’s age could potentially acquire in each language. Thus, we might expect that Jamie could potentially have learned up to 245 English words and 245 French words, although he has thus far only learned 50 in English and 20 in French.

@@

## Current research
The current research aimed to better understand the nature of translation equivalent learning in bilingual children. Study 1 simulated the expected patterns of translation equivalent learning under the Bilingual Vocabulary Model proposed in the introduction, with reference to the proportion of words learned in the dominant and non-dominant language and the number of words that are learnable at various developmental levels. We also compared predicted learning outcomes for when translation equivalents are harder to learn, or easier to learn, or similar to learn than singlets.

In Study 2, we examined real-world translation equivalent development in light of the predictions from the Bilingual Vocabulary Model, using archival data from 200 French–English bilingual children aged 18 to 33 months, whose vocabularies and translation equivalent knowledge were measured by parent report using the MacArthur-Bates CDI: Words and Sentences form in English [@Fenson_etal_2007] and Québec French (Trudeau et al., 1999). Together, the Bilingual Vocabulary Model and real-world data allowed us to examine contrasting hypotheses about translation equivalents: whether translation equivalents learning is harder (Avoidance Account), whether translation equivalent learning is easier (Preference Account), or similar to learn than singlets (Neutral Account).

@@

Translation equivalent (TE) knowledge was first simulated across different developmental levels (as indicated by number of LEARNABLE words = 150, 300, 450, 600), in conjunction with a wide range of values for DOM and NONDOM. Again, BALANCE and word vocabulary (WORD) were calculated based on the values of DOM and NONDOM. The final simulated data set contained 166 data points (see Table 2 for a summary of the parameters used). Three scenarios of translation equivalent learning (TE) were then generated using the formula TE = BIAS × DOM×NONDOM/LEARNABLE. To illustrate the Avoidance Account, BIAS was set at .5 (i.e., TEs are 50% less likely to be learned than singlets). To illustrate the Neutral Account, BIAS was set at 1 (i.e., TEs are equal to learn as singlets). Finally, to illustrate the Preference Account, BIAS was set at 1.5 (i.e., TE are 50% more likely to be learned than singlets). In Figure 4, we illustrate the three different scenarios of simulated translation equivalent (TE) knowledge. Again, we continue to observe a pattern consistent with prediction 1a where, in all cases, simulated children with more balanced vocabularies (BALANCE) produced more translation equivalents (TE). Thus, overall relationships between BALANCE and TE remained similar across the Avoidance, Preference, and Neutral Accounts. What changed was the slope of translation equivalent learning: the slopes were the shallowest under the Avoidance Account where BIAS = 0.5, whereas the slopes were steepest under the Preference Account where BIAS = 1.5. With this, we further outline Prediction Set 3:

* Prediction 3: Whether translation equivalents are harder to learn, easier to learn, or similar to learn than singlets will change the slope of translation equivalent learning as a function of word vocabulary (WORD), with a shallower slope if TEs are less easily learned (i.e., Avoidance Account), and a steeper slope if TEs are more easily learned (i.e., Preference Account) compared to where translation equivalents are similar to learn as singlets (i.e., Neutral Account).

@@

## 2.1.3 Procedure
Caregivers were asked to fill out the CDI questionnaires as part of their child’s participation in experimental studies on language development, speech perception, and word learning. Caregivers were instructed to check off the words produced by their child using either a CDI paper questionnaire or the same questionnaire administered on a tablet. Data from paper based questionnaires was double entered and checked by trained research assistants.

@@

Children produced an average of `r vocabulary_average$mean[2]` translation equivalents (TE; SD = `r vocabulary_average$sd[2]`, range = `r vocabulary_average$min[2]` - `r vocabulary_average$max[2]`). The remainder of words were singlets: Children produced many more singlets in their dominant language (DOM-SINGLET; M = `r vocabulary_average$mean[3]`, SD = `r vocabulary_average$sd[3]`, range = `r vocabulary_average$min[3]` - `r vocabulary_average$max[3]`) than in their non-dominant language (NONDOM-SINGLET; M = `r vocabulary_average$mean[4]`, SD = `r vocabulary_average$sd[4]`, range = `r vocabulary_average$min[4]` - `r vocabulary_average$max[4]`), _t_(`r t_test_SINGLET$parameter[1]`) = `r t_test_SINGLET$statistic[1]`, $p `r papaja::printp(t_test_SINGLET$p.value[1])`$, $d = `r round(d_SINGLET, 2)`$. On average, children’s concept vocabulary size was `r vocabulary_average$mean[1]` (CONCEPT; SD = `r vocabulary_average$sd[1]`, range = `r vocabulary_average$min[1]` - `r vocabulary_average$max[1]`).

@@

Prediction 2a was that older children (i.e., those at a later developmental level) would have larger concept vocabularies than younger children (i.e., those at an earlier developmental level). We observed a positive correlation between age (used as a proxy for developmental level, which determines LEARNABLE) and concept vocabulary (CONCEPT) in our dataset, _r_(`r cor_age_CONCEPT$parameter[1]`) = `r round(cor_age_CONCEPT$estimate[1],2)`, $p `r papaja::printp(cor_age_CONCEPT$p.value[1])`$, and therefore confirmed the prediction. This pattern was further confirmed by a one-way ANOVA, where the three age groups significantly differed in the number of concept vocabulary they produced, ($F$(`r aov_age_CONCEPT$df[1]`, `r aov_age_CONCEPT$df[2]`) = `r round(aov_age_CONCEPT$statistic[1],2)`, $p `r papaja::printp(aov_age_CONCEPT$p.value[1])`$). Older children of 28–33 months (i.e., at a later developmental level) produced the most with an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[3],2)` ($p$s < .001), those middle children of 23–27 months (i.e., at an intermediate developmental level) produced an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[2],2)`, and those younger children of 18–22 months (i.e., at an earlier developmental level) produced the least with an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[1],2)` ($p$s < .001).

@@

## 2.2.4 Testing Prediction Set 3: Rate of translation equivalent learning
Prediction Set 3 pertained to the overall nature of translation equivalent learning, describing expected patterns of translation equivalent learning under the Neutral Account, the Avoidance Account, or the Preference Account. To directly test the correspondence of our data with these different accounts, we built a linear regression model predicting the observed number of translation equivalents from the Bilingual Vocabulary Model using the formula TE = DOM×NONDOM/LEARNABLE , and we allowed the model to estimate BIAS parameter.

First, we will walk through the parameters in this model. The size of dominant vocabulary (DOM) and size of non-dominant vocabulary (NONDOM) were taken to be the number of words produced by individual children observed in the vocabulary data. As for the number of learnable vocabulary (LEARNABLE), this was determined by the averaging of English and French productive CDI vocabulary at the 90th percentile at different ages which was obtained from Wordbank (Frank et al., 2016), and Table 7 lists the denominator at different ages. For example, for an 18 month-old infant, the denominator was 244.9 words which was calculated by averaging the 268.7 English words and 221.1 French words, based on what 18-month-old children would typically produce at the 90th percentile. For children who were between 31 to 33 months in our dataset, the 90th percentile of 30-month-old children was used since the 90th percentile information was available only up to 30 months.

@@

Furthermore, the intercept of the linear regression model was set at 0 since no translation equivalents are expected to be produced if a child does not know any dominant or non-dominant vocabulary (i.e., when the predictor variables are 0). To reproduce the Bilingual Vocabulary Model’s formula TE = DOM×NONDOM/LEARNABLE, an interaction between dominant and non-dominant vocabulary was entered in the model, but main effects were not included in the model (denoted in R by using a colon rather than an asterisk between the interacting predictors). Therefore, our final linear regression model equation was:

_ObservedTE ~ 0 + Dominant vocabulary:Non-dominant vocabulary/90 percentile of CDI items._

@@

_ObservedTE * 90 percentile of CDI items ~ 0 + Dominant vocabulary:Non-dominant vocabulary_)

\noindent With the observed number of translation equivalents as the dependent variable, the regression coefficient estimated by the model would indicate how the BIAS parameter was consistent with the empirical vocabulary data, which would then indicate whether bilingual children were biased towards or against learning translation equivalents. If the coefficient is close to 1, then there is no bias and translation equivalents are learned equally to other words (i.e., the Neutral Account). Otherwise, a coefficient less than 1 represents a bias against learning translation equivalents where translation equivalents are less easily learned (i.e., the Avoidance Account), and a coefficient greater than 1 represents a bias towards learning translation equivalents where translation equivalents are more easily learned (i.e., the Preference Account).

>>>>

author: 
  - name          : "Rachel Ka-Ying Tsui"
    affiliation   : "1"
    corresponding : yes    
    address       : "Department of Psychology, 7141 Sherbrooke St. West, Montreal, QC, Canada, H2T1V2.  Current affilitation: Laboratory for Language Development, RIKEN Center for Brain Science, 2-1 Hirosawa, Wako-shi, Saitama, Japan, 351-0198"
    email         : "rachelkytsui@gmail.com"
  - name          : "Ana Maria Gonzalez-Barrero"
    affiliation   : "1" 
  - name          : "Esther Schott"
    affiliation   : "1" 
  - name          : "Krista Byers-Heinlein"
    affiliation   : "1"

@@

Bilingual children must learn words that take a different form in each of their languages, but share the same or highly similar meanings. For instance, to refer to the same crisp red-skinned fruit, an English–French bilingual child must use the word “apple” when speaking English, and the word _“pomme”_ when speaking French. These cross-language synonyms are known as translation equivalents (also called doublets; Umbel et al., 1992), and are observed amongst bilingual children’s first words (e.g., David & Wei, 2008; De Houwer et al., 2006; Pearson et al., 1995). Translation equivalents are thought to hold a special status in a bilingual’s developing lexicon due to the strong overlap in their semantics. For example, studies with bilingual toddlers show that the associative semantic properties of a word in one language facilitate the activation of its translation equivalent (e.g., Bilson et al., 2015; Floccia et al., 2020; Jardak & Byers-Heinlein, 2019). That is, upon hearing the English word “apple”, the corresponding French word _“pomme”_ is more easily activated in bilinguals’ minds. In vocabulary acquisition, bilingual children must learn a first label for a referent (a “singlet”; Umbel et al., 1992) before they can learn its translation equivalent. Is translation equivalent learning different from singlet learning? The current paper contrasts three competing accounts: 1) translation equivalents are harder to learn than singlets (Avoidance Account), 2) translation equivalents are easier to learn than singlets (Preference Account), and 3) translation equivalents are similar to learn as singlets (Neutral Account). To adjudicate between these accounts, we introduce the Bilingual Vocabulary Model, which provides a computational account of vocabulary learning, with parameters including bilinguals’ vocabulary in each language and their developmental level. In Study 1, we use the Bilingual Vocabulary Model to derive a set of predictions, which we then test against vocabulary data from 200 18- to 33-month-old bilingual children in Study 2.

@@

Other evidence suggesting translation equivalents might be easier to learn than singlets comes from a study that included vocabulary-checklist data from 254 monolingual and 181 bilingual children aged 6 months to 7.5 years (Bilson et al., 2015). The researchers used a network analysis approach to investigate how translation equivalents are learned, focusing on the semantic relationships between the words (e.g., words like “cat” and “dog” are strongly semantically related). Using a statistical model that allowed free semantic relations among vocabulary data from monolingual and bilingual children, the results suggested that words were learned faster when they were semantically connected to more known words in children’s lexicons. This effect applied not only to words within the same language, but also to words across languages including translation equivalents (e.g., English “dog” and French _“chien”_) and words that had other cross-language relations (e.g., “cat” and _“chien”_). The authors then simulated bilingual vocabulaires by modeling bilingual lexicons as combinations of two independent vocabulary-size-matched monolinguals. Comparison with actual bilingual children’s vocabulary data revealed that bilingual children acquired more translation equivalents than predicted by the simulation. The authors therefore concluded that bilingual children learn translation equivalents more easily than singlets. Note that, in their study, expected translation equivalent knowledge was simulated based on the number of lexical items that overlapped between two randomly chosen English monolinguals (e.g., whether both monolinguals knew the word “cat”). However, it is unclear whether this is an appropriate point of comparison for bilingual children as this approach may overlook variables that impact bilinguals’ vocabulary learning, including vocabulary size in each language and the developmental level of a child — a point that we will return to later in the introduction.

@@

### Neutral Account: Translation equivalents are similar to learn as singlets
The previous accounts rely on the idea that bilingual vocabulary development unfolds differently than monolingual development, as monolinguals encounter only singlets but bilinguals encounter both singlets and translation equivalents. There is an underlying assumption that translation equivalent learning is somehow special relative to singlet learning — the Avoidance Account proposes that translation equivalents are harder to learn than singlets, whereas the Preference Account proposes that translation equivalents are easier to learn than singlets. However, it is also possible that translation equivalents are neither harder nor easier for bilingual children to learn than singlets. We call this the Neutral Account.

The Neutral Account implies that bilingual children’s two languages develop relatively independently. Indeed, language and processing measures for bilingual children tend to be tightly correlated within a particular language, and weakly if at all correlated across languages. For example, 30-month-old bilingual children’s processing efficiency in a particular language closely correlated with vocabulary size in that language, but was unrelated to vocabulary size in their other language (Marchman et al., 2010). Due to differences in the amount of language exposure, bilingual children seldom show equal vocabulary growth in both of their languages (e.g., Pearson & Fernandez, 1994; Pearson et al., 1997), and the amount of exposure to a particular language has been reported to modulate the within-language association between language processing ability and vocabulary size (Hurtado et al., 2014). Bilingual children with greater exposure to a particular language tended to process that language faster, and in turn learned more words in that language.

@@

### Vocabulary size in bilinguals’ two languages
Balance between the two vocabulary sizes is a function of the number of words bilingual children produce in each language, which tends to be tightly linked to their exposure to each language. In general, more language exposure leads to larger vocabulary size (e.g., Barnes & Garcia, 2013; Boyce et al., 2013; Hurtado et al., 2014; Marchman et al., 2010; Pearson et al., 1997; Place & Hoff, 2011). Bilingual children usually know more words in the language in which they have greater exposure (i.e; dominant language) relative to the language in which they have less exposure (i.e., non-dominant language; Pearson et al., 1997; Place & Hoff, 2011). This is because the more often a bilingual hears a language, the more opportunities there will be for learning new words in that language.

Because translation equivalents are words from different languages that refer to the same concept, the number of words a bilingual knows in each of their languages will necessarily constrain the number of translation equivalent pairs they could possibly know. For example, a child with a less balanced vocabulary across the two languages might only say 5 words in one language but many more words in the other language; this means that the child could only produce a maximum of 5 translation equivalents, regardless of how many words they know in their other language. Conversely, it seems reasonable to expect that if a child knows a similar number of words in each language and thus has a more balanced vocabulary across the two languages, there would be more potential for some of those words to be translation equivalents.

@@

\newpage
```{r box1, results="asis", tab.cap = NULL}
box1 <- data.frame(title = "To test the independence of two events from the same population, as an example, we might ask whether Psychology majors are more likely to be left-handed. To determine independence, we must know the probability of being a psychology major, the probability of being left handed, and the probability of being both a psychology major and left-handed. The numerator of these probabilities will be the number of students who are left-handed, psychology majors, and both (respectively) and the denominator will be the total population of students we are observing. Imagine a college of 1000 students. If 100/1000 (or 1/10) students are left-handed, and 200/1000 (or 1/5) students are psychology majors, then if these variables are independent we expect a proportion of 1/10 × 1/5 = 1/50 students to be left-handed psychology majors. To determine the number we expect to observe in the college, we multiply 1/50 × 1000 = 20 students. When we compare this expected number to the actual number of left-handed students, there are three possible outcomes. First, we may observe many more than 20 left-handed psychology students at the college (say 100 students), which suggests that being left-handed increases the probability of majoring in psychology. Or, second, we may observe many fewer than 20 left-handed psychology students at the college (say 5 students), which suggests that being left-handed decreases the probability of majoring in psychology (in this example by a factor of 1/4). Finally, if left-handedness and majoring in psychology are independent, we can predict the number of left-handed psychology students by multiplying the observed number of left-handed students (100) by the observed number of psychology students (200), and dividing by the total population of the college (1000), so for example 100×200/1000 = 20. Thus, comparing expected and observed numbers can inform us about the independence of the underlying phenomena.") %>%
  rename("Box 1. Example of test for independence." = title)

@@

The model takes four main parameters: the number of words produced in the dominant language (DOM), the number of words produced in the non-dominant language (NONDOM), vocabulary size of potentially learnable words in each language (LEARNABLE), and a bias parameter (BIAS) which indicates whether the model is biased towards (BIAS > 1) or against (BIAS < 1) learning translation equivalents. The language in which a child knows more words is the dominant language, whereas the one in which a child knows fewer words is the non-dominant language. 
Next, we turn to the LEARNABLE parameter (i.e., the number of potentially learnable words). If DOM and NONDOM are measured with an instrument such as the MacArthur-Bates Communicative Development inventories (CDI; Fenson et al., 2007), one option would be to set LEARNABLE to be the total number of items on the CDI. For convenience, consider the effect of setting LEARNABLE to 600, as a round number (the actual number of CDI items is usually slightly higher than 600, depending on the language of the adaptation). Very young children would not be expected to know many of the “harder” words on the CDI, such as “lawn mower”, “sidewalk”, or “vitamins", due to children’s immature cognitive machinery and conceptual development. A more reasonable solution might be to determine how many CDI words are potentially learnable given the child’s developmental level, which could be approximated by their age. For example, imagine that Jamie who is 18 months old produces 50 English words and 20 French words, thus a total of 70 words. Monolingual children his age with the very largest productive vocabularies (those at the 90th percentile averaging between English and French norms) produce a total of 245 words (retrieved from the Wordbank database; Frank et al., 2016). Although there is likely considerable individual variability as to the cognitive capacity even amongst children of the same age, we argue that this provides a reasonable — if imperfect — estimate of the number of learnable words (LEARNABLE) that a child of Jamie’s age could potentially acquire in each language. Thus, we might expect that Jamie could potentially have learned up to 245 English words and 245 French words, although he has thus far only learned 50 in English and 20 in French.

@@

## Current research
The current research aimed to better understand the nature of translation equivalent learning in bilingual children. Study 1 simulated the expected patterns of translation equivalent learning under the Bilingual Vocabulary Model proposed in the introduction, with reference to the proportion of words learned in the dominant and non-dominant language and the number of words that are learnable at various developmental levels. We also compared predicted learning outcomes for when translation equivalents are harder, easier, or similar to learn as singlets.

In Study 2, we examined real-world translation equivalent development in light of the predictions from the Bilingual Vocabulary Model, using archival data from 200 French–English bilingual children aged 18 to 33 months, whose vocabularies and translation equivalent knowledge were measured by parent report using the MacArthur-Bates CDI: Words and Sentences form in English [@Fenson_etal_2007] and Québec French (Trudeau et al., 1999). Together, the Bilingual Vocabulary Model and real-world data allowed us to examine contrasting hypotheses about translation equivalents: whether translation equivalents learning is harder (Avoidance Account), easier (Preference Account), or similar (Neutral Account) to learning singlets.

@@

Translation equivalent (TE) knowledge was first simulated across different developmental levels (as indicated by number of LEARNABLE words = 150, 300, 450, 600), in conjunction with a wide range of values for DOM and NONDOM. Again, BALANCE and word vocabulary (WORD) were calculated based on the values of DOM and NONDOM. The final simulated data set contained 166 data points (see Table 2 for a summary of the parameters used). Three scenarios of translation equivalent learning (TE) were then generated using the formula TE = BIAS × DOM×NONDOM/LEARNABLE. To illustrate the Avoidance Account, BIAS was set at .5 (i.e., TEs are 50% less likely to be learned than singlets). To illustrate the Neutral Account, BIAS was set at 1 (i.e., TEs are equally learnable as singlets). Finally, to illustrate the Preference Account, BIAS was set at 1.5 (i.e., TE are 50% more likely to be learned than singlets). In Figure 4, we illustrate the three different scenarios of simulated translation equivalent (TE) knowledge. Again, we continue to observe a pattern consistent with prediction 1a where, in all cases, simulated children with more balanced vocabularies (BALANCE) produced more translation equivalents (TE). Thus, overall relationships between BALANCE and TE remained similar across the Avoidance, Preference, and Neutral Accounts. What changed was the slope of translation equivalent learning: the slopes were the shallowest under the Avoidance Account where BIAS = 0.5, whereas the slopes were steepest under the Preference Account where BIAS = 1.5. With this, we further outline Prediction Set 3:

* Prediction 3: Whether translation equivalents are harder to learn, easier to learn, or similar to learn as singlets will change the slope of translation equivalent learning as a function of word vocabulary (WORD), with a shallower slope if TEs are less easily learned (i.e., Avoidance Account), and a steeper slope if TEs are more easily learned (i.e., Preference Account) compared to where translation equivalents are similar to learn as singlets (i.e., Neutral Account).

@@

## 2.1.3 Procedure
Caregivers were asked to fill out the CDI questionnaires as part of their child’s participation in experimental studies on language development, speech perception, and word learning. Caregivers were instructed to check off the words produced by their child using either a CDI paper questionnaire or the same questionnaire administered on a tablet. Data from paper based questionnaires were double entered and checked by trained research assistants.

@@

Children produced an average of `r vocabulary_average$mean[2]` translation equivalents (TE; SD = `r vocabulary_average$sd[2]`, range = `r vocabulary_average$min[2]` - `r vocabulary_average$max[2]`). The remainder of words were singlets, and children produced many more singlets in their dominant language (DOM-SINGLET; M = `r vocabulary_average$mean[3]`, SD = `r vocabulary_average$sd[3]`, range = `r vocabulary_average$min[3]` - `r vocabulary_average$max[3]`) than in their non-dominant language (NONDOM-SINGLET; M = `r vocabulary_average$mean[4]`, SD = `r vocabulary_average$sd[4]`, range = `r vocabulary_average$min[4]` - `r vocabulary_average$max[4]`), _t_(`r t_test_SINGLET$parameter[1]`) = `r t_test_SINGLET$statistic[1]`, $p `r papaja::printp(t_test_SINGLET$p.value[1])`$, $d = `r round(d_SINGLET, 2)`$. On average, children’s concept vocabulary size was `r vocabulary_average$mean[1]` (CONCEPT; SD = `r vocabulary_average$sd[1]`, range = `r vocabulary_average$min[1]` - `r vocabulary_average$max[1]`).

@@

Prediction 2a was that older children (i.e., those at a later developmental level) would have larger concept vocabularies than younger children (i.e., those at an earlier developmental level). We observed a positive correlation between age (used as a proxy for developmental level, which determines LEARNABLE) and concept vocabulary (CONCEPT) in our dataset, _r_(`r cor_age_CONCEPT$parameter[1]`) = `r round(cor_age_CONCEPT$estimate[1],2)`, $p `r papaja::printp(cor_age_CONCEPT$p.value[1])`$, and therefore confirmed the prediction. This pattern was further confirmed by a one-way ANOVA, where the three age groups significantly differed in the number of concept vocabulary they produced, ($F$(`r aov_age_CONCEPT$df[1]`, `r aov_age_CONCEPT$df[2]`) = `r round(aov_age_CONCEPT$statistic[1],2)`, $p `r papaja::printp(aov_age_CONCEPT$p.value[1])`$). Older children of 28–33 months (i.e., at a later developmental level) produced the most with an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[3],2)` ($p$s < .001), middle children of 23–27 months (i.e., at an intermediate developmental level) produced an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[2],2)`, and younger children of 18–22 months (i.e., at an earlier developmental level) produced the least with an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[1],2)` ($p$s < .001).

@@

## 2.2.4 Testing Prediction Set 3: Rate of translation equivalent learning
Prediction Set 3 pertained to the overall nature of translation equivalent learning, describing expected patterns of translation equivalent learning under the Neutral Account, the Avoidance Account, or the Preference Account. To directly test the correspondence of our data with these different accounts, we built a linear regression model predicting the observed number of translation equivalents from the Bilingual Vocabulary Model using the formula TE = DOM×NONDOM/LEARNABLE , and we allowed the model to estimate the BIAS parameter.

First, we will walk through the parameters in this model. The size of dominant vocabulary (DOM) and size of non-dominant vocabulary (NONDOM) were taken to be the number of words produced by individual children observed in the vocabulary data. The number of learnable words (LEARNABLE) was determined by averaging English and French productive CDI vocabulary at the 90th percentile at different ages (percentile information obtained from Wordbank; Frank et al., 2016). Table 7 lists the denominators derived from averaging these percentiles, or LEARNABLE, at different ages. For example, for an 18 month-old infant, the denominator was 244.9 words which was calculated by averaging the 268.7 English words and 221.1 French words, based on what 18-month-old children would typically produce at the 90th percentile. For children who were between 31 to 33 months in our dataset, the 90th percentile of 30-month-old children was used since 90th percentile information was available only up to 30 months.

@@

Furthermore, the intercept of the linear regression model was set at 0 since no translation equivalents are expected to be produced if a child does not know any dominant or non-dominant vocabulary (i.e., when the predictor variables are 0). To reproduce the Bilingual Vocabulary Model’s formula TE = DOM×NONDOM/LEARNABLE, an interaction between dominant and non-dominant vocabulary was entered in the model, but main effects were not included (denoted in R by using a colon rather than an asterisk between the interacting predictors). Therefore, our final linear regression model equation was:

_ObservedTE ~ 0 + Dominant vocabulary:Non-dominant vocabulary/90th percentile of CDI items._

@@

_ObservedTE * 90th percentile of CDI items ~ 0 + Dominant vocabulary:Non-dominant vocabulary_)

\noindent With the observed number of translation equivalents as the dependent variable, the regression coefficient estimated by the model would indicate how the BIAS parameter was related to the empirical vocabulary data, which would then indicate whether bilingual children were biased towards or against learning translation equivalents. If the coefficient is close to 1, then there is no bias and translation equivalents are learned equally to other words (i.e., the Neutral Account). Otherwise, a coefficient less than 1 represents a bias against learning translation equivalents where translation equivalents are less easily learned (i.e., the Avoidance Account), and a coefficient greater than 1 represents a bias towards learning translation equivalents where translation equivalents are more easily learned (i.e., the Preference Account).
